---
title: "READMITTED"
subtitle: "How America's Hospitals Are Failing the 30-Day Standard"
format:
  html:
    number-sections: false
    theme: none
    css: artometrics.css
    toc: true
    toc-depth: 2
    toc-title: "IN THIS REPORT"
    toc-location: left
    smooth-scroll: true
    embed-resources: true
    code-fold: true
    code-tools: true
editor:
  markdown:
    wrap: 72
---

```{r setup}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(httr)
library(jsonlite)
library(scales)
library(ggtext)
library(here)

# ── Artometrics color palette ──────────────────────────────────────
art_bg        <- "#FAFAF8"
art_cream     <- "#F2F0EB"
art_dark      <- "#1C1C1E"
art_mid       <- "#6B6B6B"
art_highlight <- "#C0392B"
art_secondary <- "#2C3E6B"
art_muted     <- "#D5D5D5"

# ── Artometrics ggplot2 theme ──────────────────────────────────────
theme_artometrics <- function() {
  theme_minimal(base_family = "Helvetica") +
    theme(
      plot.background    = element_rect(fill = art_cream, color = NA),
      panel.background   = element_rect(fill = art_cream, color = NA),
      panel.grid.major   = element_line(color = art_muted, linewidth = 0.4),
      panel.grid.minor   = element_blank(),
      axis.text          = element_text(color = art_mid, size = 9),
      axis.title         = element_text(color = art_dark, size = 10,
                                        face = "bold"),
      plot.title         = element_markdown(color = art_dark, size = 13,
                                            face = "bold",
                                            margin = margin(b = 6)),
      plot.subtitle      = element_markdown(color = art_mid, size = 10,
                                            margin = margin(b = 10)),
      plot.caption       = element_text(color = art_mid, size = 8,
                                        hjust = 0, margin = margin(t = 10)),
      plot.caption.position = "plot",
      plot.margin        = margin(t = 16, r = 16, b = 16, l = 16),
      legend.background  = element_rect(fill = art_cream, color = NA),
      legend.text        = element_text(color = art_mid, size = 8),
      legend.title       = element_text(color = art_dark, size = 9,
                                        face = "bold"),
      strip.text         = element_text(color = art_dark, size = 9,
                                        face = "bold"),
      strip.background   = element_rect(fill = art_cream, color = NA)
    )
}
```

```{r load-data}
#| echo: false
#| warning: false
#| message: false

meta_url <- paste0(
  "https://data.cms.gov/provider-data/api/1/metastore/",
  "schemas/dataset/items/9n3s-kdb3?show-reference-ids=false"
)

meta         <- fromJSON(content(GET(meta_url), "text", encoding = "UTF-8"))
download_url <- meta$distribution$data$downloadURL
hrrp         <- read_csv(download_url, na = c("N/A", "", " "),
                         show_col_types = FALSE)
```

```{r clean-data}
#| echo: false
#| warning: false
#| message: false

hrrp_clean <- hrrp |>
  rename(
    facility_name  = `Facility Name`,
    facility_id    = `Facility ID`,
    state          = State,
    measure_name   = `Measure Name`,
    discharges     = `Number of Discharges`,
    footnote       = Footnote,
    err            = `Excess Readmission Ratio`,
    predicted_rate = `Predicted Readmission Rate`,
    expected_rate  = `Expected Readmission Rate`,
    readmissions   = `Number of Readmissions`,
    start_date     = `Start Date`,
    end_date       = `End Date`
  ) |>
  filter(!is.na(err)) |>
  mutate(
    condition = case_when(
      measure_name == "READM-30-AMI-HRRP"      ~ "AMI",
      measure_name == "READM-30-HF-HRRP"       ~ "Heart Failure",
      measure_name == "READM-30-PN-HRRP"       ~ "Pneumonia",
      measure_name == "READM-30-COPD-HRRP"     ~ "COPD",
      measure_name == "READM-30-HIP-KNEE-HRRP" ~ "Hip/Knee",
      measure_name == "READM-30-CABG-HRRP"     ~ "CABG",
      TRUE ~ measure_name
    ),
    penalty_tier = case_when(
      err <= 1.000 ~ "No Penalty",
      err <= 1.010 ~ "Low",
      err <= 1.025 ~ "Medium",
      err >  1.025 ~ "High"
    ),
    penalty_tier = factor(penalty_tier,
                          levels = c("No Penalty", "Low", "Medium", "High"))
  )
```

```{r fast-facts-compute}
#| echo: false
#| warning: false
#| message: false
#| results: hide

# All values computed silently — referenced in Fast Facts HTML below.

pct_penalized <- mean(hrrp_clean$err > 1.0) * 100  # 48.1

avg_err <- mean(hrrp_clean$err)                     # 1.0018

err_by_condition <- hrrp_clean |>
  group_by(condition) |>
  summarise(avg_err = mean(err)) |>
  arrange(desc(avg_err)) |>
  as.data.frame()
# 1 Hip/Knee  1.003985
# 2 CABG      1.001789
# 3 AMI       1.001755
# 4 Pneumonia 1.001486
# 5 HF        1.001439
# 6 COPD      1.001133

worst_state <- hrrp_clean |>
  group_by(state) |>
  summarise(
    total     = n(),
    penalized = sum(err > 1.0),
    pct       = penalized / total * 100
  ) |>
  arrange(desc(pct)) |>
  slice(1)
# NJ — 65.4%
```

## FAST FACTS

```{=html}
<style>
  .facts-grid {
    display: grid;
    grid-template-columns: repeat(3, 1fr);
    gap: 16px;
    margin: 24px 0 32px 0;
  }
  .fact-box {
    background: #F2F0EB;
    border-left: 4px solid #C0392B;
    padding: 18px 20px;
    border-radius: 2px;
    text-align: left;
  }
  .fact-number {
    display: block;
    font-size: 1.75rem;
    font-weight: 700;
    color: #C0392B;
    line-height: 1.1;
    margin-bottom: 6px;
    font-family: Helvetica, sans-serif;
    text-align: left;
  }
  .fact-label {
    display: block;
    font-size: 0.82rem;
    color: #6B6B6B;
    line-height: 1.45;
    font-family: Helvetica, sans-serif;
    text-align: left;
    max-width: none;
    margin: 0;
  }
  @media (max-width: 700px) {
    .facts-grid { grid-template-columns: 1fr 1fr; }
  }
</style>

<div class="facts-grid">
  <div class="fact-box">
    <span class="fact-number">48.1%</span>
    <span class="fact-label">of HRRP-eligible hospital-condition
    pairs carried an excess readmission ratio above 1.0 —
    more readmissions than CMS models predict</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">6</span>
    <span class="fact-label">conditions tracked under HRRP —
    AMI, Heart Failure, Pneumonia, COPD, Hip/Knee, and CABG</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">1.0018</span>
    <span class="fact-label">average excess readmission ratio
    nationally — the average hospital readmits slightly more
    patients than CMS models expect</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">Hip/Knee</span>
    <span class="fact-label">the highest-ERR condition on average
    (1.004) — joint replacement patients are the hardest
    readmission problem for hospitals to control</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">742</span>
    <span class="fact-label">hospitals mandated under the CMS TEAM
    model as of January 1, 2026 — Rainfall Health's
    primary market</span>
  </div>
  <div class="fact-box">
    <span class="fact-number">NJ</span>
    <span class="fact-label">state with the highest share of
    penalized hospital-condition pairs at 65.4% —
    the highest-risk geography in the dataset</span>
  </div>
</div>
```

### WHAT HRRP ACTUALLY IS

The Hospital Readmissions Reduction Program is a Medicare penalty system that has been running since 2012. The premise is simple: if your patients come back to the hospital within 30 days of discharge at a higher rate than expected given their medical profile, CMS docks your Medicare reimbursements. The penalty caps at 3% of all Medicare payments — not just payments for the specific condition being measured. A hospital with too many heart failure readmissions doesn't just lose heart failure revenue. It loses 3% of everything.

This creates an interesting dynamic. The penalty is universal but the measurement is condition-specific. CMS tracks six conditions: Acute Myocardial Infarction (AMI), Heart Failure (HF), Pneumonia, Chronic Obstructive Pulmonary Disease (COPD), Hip & Knee Arthroplasty, and Coronary Artery Bypass Graft (CABG). A hospital can perform flawlessly on five of them and still get penalized because of one.

### HOW THE EXCESS READMISSION RATIO WORKS

The core metric in this dataset is the Excess Readmission Ratio, or ERR. It is a ratio of predicted readmissions to expected readmissions — where "expected" is risk-adjusted for each patient's age, comorbidities, and discharge history. An ERR of exactly 1.0 means you readmitted exactly as many patients as a hospital of your type and patient mix should. Above 1.0 means more readmissions than expected. Below 1.0 means fewer.

What this analysis found: nearly every condition's national average ERR is above 1.0. Not by a lot — the spread is in the fourth decimal place — but it's consistently above. HRRP has been running for over a decade, and the average American hospital is still readmitting more patients than CMS expects. The penalty hasn't eliminated the problem. It has clarified it.

### WHAT THE DATA COVERS — AND WHAT IT DOESN'T

The FY2024 HRRP dataset covers the measurement period July 1, 2021 through June 30, 2024. After removing suppressed rows — hospitals with fewer than 25 discharges for a given condition, which CMS redacts to protect patient privacy — the working dataset contains 11,720 hospital-condition pairs across roughly 2,700 hospitals and 6 conditions.

The suppressed rows matter. Small rural hospitals disproportionately fall below the 25-discharge threshold and disappear from the analysis. The hospitals you see in this data skew toward larger, busier facilities. That's not a flaw in the data — it's a feature of how CMS designed the program. But it's context that shapes every finding below.

### THE PENALTY IS THE FLOOR, NOT THE CEILING

An ERR above 1.0 triggers a penalty calculation, but the penalty tier depends on how far above 1.0 you are and the hospital's overall performance relative to its peer group. For this analysis, hospitals are grouped into four tiers: No Penalty (ERR ≤ 1.0), Low (just above), Medium, and High. The High tier — hospitals with the worst readmission performance — is where CMS's new TEAM model comes in. Starting January 1, 2026, 742 hospitals are mandated to participate in TEAM, a bundled payment model that goes further than HRRP by tying entire episodes of care — not just readmissions — to reimbursement. HRRP is where the penalty starts. TEAM is where the stakes get real.


## SQL EXHIBIT — HRRP CLEANING LOGIC

The following query mirrors the R cleaning and classification logic used in this report. It is included as a display-only portfolio piece demonstrating SQL literacy alongside the R analysis above.

::: {.sql-exhibit}
```sql
-- HRRP READMISSIONS: CLEANING AND PENALTY TIER CLASSIFICATION
-- Mirrors the R cleaning logic in readmitted.qmd
-- Source: CMS Provider Data Catalog, Dataset ID: 9n3s-kdb3

SELECT
    facility_name,
    facility_id,
    state,
    measure_name,

    -- Core readmission metrics
    CAST(excess_readmission_ratio AS FLOAT)     AS err,
    CAST(number_of_discharges AS INT)           AS discharges,
    CAST(predicted_readmission_rate AS FLOAT)   AS predicted_rate,
    CAST(expected_readmission_rate AS FLOAT)    AS expected_rate,

    -- Penalty tier classification
    CASE
        WHEN CAST(excess_readmission_ratio AS FLOAT) <= 1.0
            THEN 'No Penalty'
        WHEN CAST(excess_readmission_ratio AS FLOAT) <= 1.005
            THEN 'Low'
        WHEN CAST(excess_readmission_ratio AS FLOAT) <= 1.015
            THEN 'Medium'
        ELSE
            'High'
    END AS penalty_tier

FROM hrrp_readmissions

WHERE
    -- Drop suppressed rows (CMS redacts hospitals < 25 discharges)
    excess_readmission_ratio NOT IN ('Too Few Cases', 'N/A', '')
    AND excess_readmission_ratio IS NOT NULL

ORDER BY
    state,
    facility_name,
    measure_name;


-- AGGREGATE VIEW: Average ERR by condition and state
SELECT
    measure_name,
    state,
    COUNT(*)                                               AS hospital_count,
    ROUND(AVG(CAST(excess_readmission_ratio AS FLOAT)), 6) AS avg_err,
    ROUND(SUM(CASE
        WHEN CAST(excess_readmission_ratio AS FLOAT) > 1.0
        THEN 1 ELSE 0
    END) * 100.0 / COUNT(*), 1)                           AS pct_penalized

FROM hrrp_readmissions

WHERE
    excess_readmission_ratio NOT IN ('Too Few Cases', 'N/A', '')
    AND excess_readmission_ratio IS NOT NULL

GROUP BY
    measure_name,
    state

ORDER BY
    avg_err DESC;
```
:::

```{r chart1_states_penalized}
library(tidyverse)
library(here)

dir.create(here("charts"), showWarnings = FALSE)

state_pct <- hrrp_clean %>%
  group_by(state) %>%
  summarise(
    pct_penalized = mean(err > 1.0, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(pct_penalized)) %>%
  slice_head(n = 20) %>%
  mutate(
    state = fct_reorder(state, pct_penalized),
    above_avg = pct_penalized > 48.1,
    bar_color = if_else(above_avg, "#C0392B", "#2C3E6B")
  )

nat_avg <- 48.1

ggplot(state_pct, aes(x = pct_penalized, y = state)) +
  geom_col(fill = state_pct$bar_color, width = 0.65) +
  geom_vline(xintercept = nat_avg, linetype = "dashed", color = "#1C1C1E", linewidth = 0.5) +
  annotate(
    "text", x = nat_avg - 0.6, y = 19.5,
    label = paste0("National avg\n", nat_avg, "%"),
    hjust = 1, vjust = 1, size = 3, color = "#6B6B6B",
    family = "sans"
  ) +
  scale_x_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Which States Have the Most Penalized Hospitals?",
    subtitle = "Top 20 states by share of hospital-condition pairs with ERR > 1.0",
    x = "% of Hospital-Condition Pairs with ERR > 1.0",
    y = NULL,
    caption = "Source: CMS Hospital Readmissions Reduction Program (HRRP) | Artometrics"
  ) +
  theme_artometrics() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "#D5D5D5", linewidth = 0.3)
  )

ggsave(here("charts", "chart1_states_penalized.png"), width = 10, height = 7, dpi = 300, bg = "#FAFAF8")
```


```{r chart2_err_by_condition}

condition_err <- hrrp_clean %>%
  group_by(condition) %>%
  summarise(
    avg_err = mean(err, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_err)) %>%
  mutate(condition = fct_reorder(condition, avg_err))

x_min <- floor(min(condition_err$avg_err) * 10000) / 10000 - 0.0005
x_max <- ceiling(max(condition_err$avg_err) * 10000) / 10000 + 0.0005

ggplot(condition_err, aes(x = avg_err, y = condition)) +
  geom_segment(
    aes(x = 1.0, xend = avg_err, yend = condition),
    color = "#D5D5D5", linewidth = 0.8
  ) +
  geom_point(color = "#C0392B", size = 4) +
  geom_vline(xintercept = 1.0, color = "#1C1C1E", linewidth = 0.6) +
  geom_text(
    aes(label = sprintf("%.4f", avg_err)),
    hjust = -0.35, size = 3.2, color = "#1C1C1E", family = "sans"
  ) +
  scale_x_continuous(
    limits = c(x_min, x_max + 0.0012),
    labels = function(x) sprintf("%.4f", x),
    expand = expansion(mult = c(0.01, 0.08))
  ) +
  annotate(
    "text", x = 1.0, y = 6.6,
    label = "ERR = 1.0\n(no excess)", hjust = 0.5, vjust = 0,
    size = 2.8, color = "#6B6B6B", family = "sans"
  ) +
  labs(
    title = "Excess Readmission Ratio by Condition",
    subtitle = "All six conditions exceed 1.0 — but the spread is measured in thousandths",
    x = "Average Excess Readmission Ratio (ERR)",
    y = NULL,
    caption = "Source: CMS HRRP | Artometrics"
  ) +
  theme_artometrics() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "#D5D5D5", linewidth = 0.3)
  )

ggsave(here("charts", "chart2_err_by_condition.png"), width = 10, height = 6, dpi = 300, bg = "#FAFAF8")

```


```{r chart3_penalty_by_ownership}
library(httr2)
library(janitor)

hospital_info <- read_csv("https://data.cms.gov/provider-data/sites/default/files/resources/893c372430d9d71a1c52737d01239d47_1770163599/Hospital_General_Information.csv") %>%
  clean_names() %>%
  select(facility_id, hospital_ownership) %>%
  mutate(
    ownership_group = case_when(
      str_detect(hospital_ownership, regex("government|tribal|district", ignore_case = TRUE)) ~ "Government",
      str_detect(hospital_ownership, regex("non-profit|voluntary|church", ignore_case = TRUE)) ~ "Non-Profit",
      str_detect(hospital_ownership, regex("proprietary|for.profit|physician", ignore_case = TRUE)) ~ "For-Profit",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(ownership_group))

joined <- hrrp_clean %>%
  inner_join(hospital_info, by = "facility_id")

tier_order <- c("No Penalty", "Low", "Medium", "High")
tier_colors <- c(
  "No Penalty" = "#2C3E6B",
  "Low"        = "#7F8FA6",
  "Medium"     = "#E07B54",
  "High"       = "#C0392B"
)

plot_data <- joined %>%
  filter(!is.na(penalty_tier)) %>%
  mutate(
    penalty_tier = factor(penalty_tier, levels = tier_order),
    ownership_group = factor(ownership_group, levels = c("Non-Profit", "Government", "For-Profit"))
  ) %>%
  count(ownership_group, penalty_tier) %>%
  group_by(ownership_group) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

ggplot(plot_data, aes(x = ownership_group, y = pct, fill = penalty_tier)) +
  geom_col(position = "fill", width = 0.6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = tier_colors, name = "Penalty Tier") +
  coord_flip() +
  labs(
    title = "Penalty Tier Distribution by Hospital Ownership Type",
    subtitle = "For-profit hospitals carry a heavier share of medium and high penalties",
    x = NULL,
    y = "Share of Hospital-Condition Pairs",
    caption = "Source: CMS HRRP + CMS Hospital General Information (xubh-q36u) | Artometrics"
  ) +
  theme_artometrics() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(color = "#D5D5D5", linewidth = 0.3),
    legend.position = "right"
  )

ggsave(here("charts", "chart3_penalty_by_ownership.png"), width = 10, height = 5, dpi = 300, bg = "#FAFAF8")
```

